---
title: "F tests for ANOVA, More Submodels"
subtitle: "(Sleuth3 Sections 5.3 and 5.4)"
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
  - \usepackage{multirow}
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

## Spock Trials Example

```{r, message = FALSE, echo = FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(mosaic)
library(gmodels)

juries <- read_csv("http://www.evanlray.com/data/sleuth3/ex0502_women_jurors.csv")
head(juries)

ggplot(data = juries, mapping = aes(x = Judge, y = Percent)) +
  geom_boxplot()
```

#### All means equal?

$H_0: \mu_1 = \mu_2 = \mu_3 = \mu_4 = \mu_5 = \mu_6 = \mu_7$

$H_A:$ At least one mean is not equal to the others.

```{r}
spock_fit_full <- lm(Percent ~ Judge, data = juries)
anova(spock_fit_full)
```

Note: Reduced model has only one mean for all 7 ($H_0$ is true).

What is the conclusion?

\vspace{3cm}

#### Means for other 6 judges equal?


$H_0: \mu_1 = \mu_2 = \mu_3 = \mu_4 = \mu_5 = \mu_6$

$H_A:$ At least one mean for a judge other than Spock is not equal to the others.

Note: these hypotheses involve judges A through F; Spock's judge is not involved in these hypothesis statements.\

```{r}
juries <- juries %>%
  mutate(
    judge_spocks = ifelse(Judge %in% c("A", "B", "C", "D", "E", "F"), "Not Spock's", "Spock's")
  )
head(juries, 20)

spock_fit_reduced <- lm(Percent ~ judge_spocks, data = juries)
anova(spock_fit_reduced, spock_fit_full)
```

What is the conclusion?
