---
title: "Chapter 11: Multiple Regression, Pairs Plots, and Added Variable Plots"
output:
  pdf_document:
    keep_tex: true
header-includes:
   - \usepackage{soul}
   - \usepackage{booktabs}
documentclass: extarticle
geometry: margin=0.6in
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
library(mosaic)
library(gridExtra)
library(carData)
Duncan <- carData::Duncan
Duncan$occupation <- rownames(Duncan)
```

## Duncan's Occupational Prestige Data

#### Intro to data

We have a data set with measurements on 45 different U.S. occupations as of 1950 (descriptions below are quotes from Fox and Weisberg, 2011):

 * `type`: Type of occupation. A factor with the following levels: `prof`, professional and managerial; `wc`, white-collar; `bc`, blue-collar.
 * `income`: Percentage of occupational incumbents in the 1950 US Census who earned $3,500 or more per year (about $36,000 in 2017 US dollars).
 * `education`: Percentage of occupational incumbents in 1950 who were high school graduates (which, were we cynical, we would say is roughly equivalent to a PhD in 2017)
 * `prestige`: Percentage of respondents in a social survey who rated the occupation as “good” or better in prestige
 
```{r}
head(Duncan)
```

References:

 * Fox, J. and Weisberg, S. (2011) An R Companion to Applied Regression, Second Edition, Sage.
 * Duncan, O. D. (1961) A socioeconomic index for all occupations. In Reiss, A. J., Jr. (Ed.) Occupations and Social Status. Free Press [Table VI-1].

Let's consider a model for occupational prestige as a function of income, education, and type of occupation.

We should always start with plots, but we're really hitting the limits of what's plottable now...

\newpage

#### Option 1: plotly

 * Formatting very similar to, but not exactly the same as, ggplot2
 * **Can't show output in pdf, only for html output or interactive use**
 * Can't be used for any more variables than we have in this example.
 * If plotly code doesn't give you what you want right away, it can be essentially impossible to fix (not a fully developed and functional package).

```{r, eval = FALSE}
library(plotly)
plot_ly(Duncan, x = ~income, y = ~education, z = ~prestige, color = ~type) %>%
  add_markers()
```

Here's a screenshot, will demo live:

\includegraphics[height=3in]{plotly_screenshot.png}

\newpage

#### Option 2: Pairs Plots

```{r}
library(GGally)
ggpairs(Duncan %>% select(-occupation))
```

Possible evidence of one outlier?  Let's fit some models and come back to that later.

\newpage

## A first model - income only explanatory variable

```{r}
lm_fit_1 <- lm(prestige ~ income, data = Duncan)
summary(lm_fit_1)
```

```{r, fig.height = 2}
ggplot(data = Duncan, mapping = aes(x = income, y = prestige)) +
  geom_point() +
  geom_smooth(method = "lm")
```

#### What is the equation of the estimated line?

\vspace{3cm}

#### What is the interpretation of the coefficient estimate for income?

\newpage

## Second Model: income and education as explanatory variables

```{r}
lm_fit_2 <- lm(prestige ~ income + education, data = Duncan)
summary(lm_fit_2)
```

#### What's the estimated equation of this model?

\vspace{3cm}

#### This can be visualized as a plane

Plotly code suppressed because it's awful.

```{r, echo = FALSE, eval = FALSE}
income_grid <- seq(from = min(Duncan$income), to = max(Duncan$income), length = 11)
education_grid <- seq(from = min(Duncan$education), to = max(Duncan$education), length = 11)
x_grid <- expand.grid(
  income = income_grid,
  education = education_grid
)

z <- predict(lm_fit_2, newdata = x_grid)
dim(z) <- c(11, 11)
z <- t(z)

plot_ly(showscale = FALSE) %>% #, color = ~type) %>%
  add_markers(data = Duncan, x = ~income, y = ~education, z = ~ prestige) %>%
  add_surface(x = ~income_grid, y = ~education_grid, z = ~ z, opacity = 0.8)
```

\includegraphics[height=3in]{plotly_ss_1plane.png}

#### What is the interpretation of the coefficient estimate for income?

\vspace{4cm}

#### Added Variables Plots

 * **After accounting for the effects of other variables in the model, what is the relationship between income and prestige?**
 * Focus on added variable plot for income: **after accounting for education**,
    * New X variable is *residuals* from a regression of income on education (what's left over in income, after accounting for education level?)
    * New Y variable is *residuals* from a regression of prestige on education (what's left over in prestige, after accounting for education level?)
    * The slope of the line describing the relationship between these residuals is the estimated coefficient for income from the model fit.

```{r, message = FALSE}
library(car)
avPlots(lm_fit_2)
```

```{r}
fit_income_education <- lm(income ~ education, data = Duncan)
fit_prestige_educaiton <- lm(prestige ~ education, data = Duncan)
av_df <- data.frame(
  income_resids = residuals(fit_income_education),
  prestige_resids = residuals(fit_prestige_educaiton)
)
fit_prestige_income_accounting_education <- lm(prestige_resids ~ income_resids, data = av_df)
summary(fit_prestige_income_accounting_education)
```

Compare to coefficient from model with income and education as explanatory variables.

\newpage

## Third Model: All 3 explanatory variables!

```{r}
lm_fit_3 <- lm(prestige ~ income + education + type, data = Duncan)
summary(lm_fit_3)
```

Plotly code suppressed because it's awful.

```{r, echo = FALSE, eval = FALSE}
income_grid <- seq(from = min(Duncan$income), to = max(Duncan$income), length = 11)
education_grid <- seq(from = min(Duncan$education), to = max(Duncan$education), length = 11)
x_grid <- expand.grid(
  income = income_grid,
  education = education_grid
)

x_grid$type <- "bc"
z_bc <- predict(lm_fit_3, newdata = x_grid)
dim(z_bc) <- c(11, 11)
z_bc <- t(z_bc)

x_grid$type <- "prof"
z_prof <- predict(lm_fit_3, newdata = x_grid)
dim(z_prof) <- c(11, 11)
z_prof <- t(z_prof)

x_grid$type <- "wc"
z_wc <- predict(lm_fit_3, newdata = x_grid)
dim(z_wc) <- c(11, 11)
z_wc <- t(z_wc)

color_bc = factor(rep("bc", length(z_bc)))
color_bc = rep(0, length(z_bc))
dim(color_bc) <- c(11, 11)
color_prof = rep(1, length(z_prof))
dim(color_prof) <- c(11, 11)
color_wc = rep(2, length(z_wc))
dim(color_wc) <- c(11, 11)
colors <- c("#66c2a5", "#fc8d62", "#8da0cb")

plot_ly(color = colors, showscale = FALSE) %>% #, color = ~type) %>%
  add_markers(data = Duncan %>% filter(type == "bc"), x = ~income, y = ~education, z = ~ prestige, color = "bc") %>% #~ type, colors = colors) %>% #8da0cb
  add_markers(data = Duncan %>% filter(type == "prof"), x = ~income, y = ~education, z = ~ prestige, color = "prof") %>% #~ type, colors = colors) %>%
  add_markers(data = Duncan %>% filter(type == "wc"), x = ~income, y = ~education, z = ~ prestige, color = "wc") %>% #~ type, colors = colors) %>%
  add_surface(x = ~income_grid, y = ~education_grid, z = ~ z_bc, color = color_bc, cmin = 0, cmax = 2, colorscale = list(c(0,1),c("#66c2a5","#66c2a5")), opacity = 0.8) %>%
  add_surface(x = ~income_grid, y = ~education_grid, z = ~ z_prof, color = color_prof, cmin = 0, cmax = 2, colorscale = list(c(0,1),c("#fc8d62","#fc8d62")), opacity = 0.8) %>%
  add_surface(x = ~income_grid, y = ~education_grid, z = ~ z_wc, color = color_wc, cmin = 0, cmax = 1, colorscale = list(c(0,1),c("#8da0cb","#8da0cb")), opacity = 0.8)
```

\includegraphics[height=3in]{plotly_ss_3planes.png}

#### What is the equation of the model fit?

\vspace{3cm}

#### What is the interpretation of the estimated coefficient for income?

\vspace{3cm}

#### Diagnostic Plots

```{r, fig.height = 1.75}
Duncan <- Duncan %>%
  mutate(
    obs_index = row_number(),
    h = hatvalues(lm_fit_3),
    studres = rstudent(lm_fit_3),
    D = cooks.distance(lm_fit_3)
  )

ggplot(data = Duncan, mapping = aes(x = obs_index, y = h)) +
  geom_point() +
  geom_hline(yintercept = 2 * 5 / nrow(Duncan)) +
  ylim(0, 1) +
  ggtitle("Leverage")

ggplot(data = Duncan, mapping = aes(x = obs_index, y = studres)) +
  geom_point() +
  ggtitle("Studentized Residuals")

ggplot(data = Duncan, mapping = aes(x = obs_index, y = D)) +
  geom_point() +
  ggtitle("Cook's Distance")
```

```{r}
obs_to_investigate <- c(6, 16, 27)
```

\newpage

```{r}
Duncan[obs_to_investigate, ]
```

```{r, fig.height=4, fig.width=4}
plot(lm_fit_3)
```

\newpage

```{r}
Duncan_minus_suspicious <- Duncan[-obs_to_investigate, ]
lm_fit_without_suspicious <- lm(prestige ~ income + education + type, data = Duncan_minus_suspicious)
summary(lm_fit_without_suspicious)
```

```{r}
Duncan_minus_minister <- Duncan[-6, ]
lm_fit_without_minister <- lm(prestige ~ income + education + type, data = Duncan_minus_minister)
summary(lm_fit_without_minister)
```

```{r}
Duncan <- Duncan %>%
  mutate(
    suspicious = ifelse(row_number() %in% obs_to_investigate, "suspicious", "seems ok")
  )

ggplot(data = Duncan, mapping = aes(x = education, y = income, color = suspicious)) +
  geom_point()
```

#### What do we say?
