---
title: "ANOVA: Transformations"
subtitle: "Sleuth3 Sections 3.5 and 5.5"
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(ggplot2)
library(readr)
library(dplyr)
```

## Context

 * Transformations can sometimes help with the following issues:
    * non-normal distributions within each group (but skewness is only a problem if it is very serious)
    * lack of equal variance for all groups
    * outliers (but usually only if this is a side effect of serious skewness)

 * The most common transformations (that we'll consider in this class) work for positive values only.

## The Ladder of Powers

 * Imagine a "ladder of powers" of $y$: We start at $y$ and go up or down the ladder.

\begin{table}[!h]
\centering
\begin{tabular}{c p{8cm}}
\toprule
Transformation & Comments \\
\toprule
\vdots & \\
\midrule
$e^y$ & Exactly where on the ladder the exponential transformation belongs depends on the magnitude of the data, but somewhere around here... \\
\midrule
$y^2$ & \\
\midrule
$y$  & Start here (no transformation) \\
\midrule
$\sqrt{y}$ & \\
\midrule
$y^{``0"}$  & We use $\log(y)$ here \\
\midrule
$-1/\sqrt{y}$ & The $-$ keeps the values of $y$ in order \\
\midrule
$-1/y$ & \\
\midrule
$-1/y^2$ & \\
\midrule
\vdots & \\
\bottomrule
\end{tabular}
\end{table}

 * Which direction?
    * If a variable is skewed right, move it down the ladder (pull down large values)
    * If a variable is skewed left, move it up the ladder (pull up small values)

## Some (minimal) facts about logarithms and exponentials

 * Foundations:
    * In this class the base of our logarithms is $e$
    * Notation: $\exp(x) = e^x$
 * $\log()$ and $\exp()$ are inverses
    * $\log(\exp(x)) = x$
    * $\exp(\log(x)) = x$
 * They are useful because they convert multiplication to addition, and addition to multiplication
     * $\log(a \cdot b) = \log(a) + \log(b)$
     * $\exp(a + b) = \exp(a) \cdot \exp(b)$

\newpage

```{r, fig.height = 1}
example <- data.frame(
  y = c(1, 2, 3, 4),
    y_squared = c(1, 2, 3, 4)^2,
    sqrt_y = c(1, 2, 3, 4)^0.5
)
example

ggplot(data = example, mapping = aes(x = y_squared, y = 0)) +
  geom_point() +
  ggtitle("Moved Up 1 Step: spread out points on the right side") +
  ylab("") +
  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank())

ggplot(data = example, mapping = aes(x = y, y = 0)) +
  geom_point() +
  ggtitle("Starting Point: evenly spaced") +
  ylab("") +
  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank())

ggplot(data = example, mapping = aes(x = sqrt_y, y = 0)) +
  geom_point() +
  ggtitle("Moved Down 1 Step: spread out points on the left side") +
  ylab("") +
  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank())
```

\newpage

## Example: Cloud Seeding (Sleuth3 Case Study 3.1.1)

Quote from book: "On each of 52 days that were deemed suitable for cloud seeding, a random mechanism was used to decide whether to seed the target cloud on that day or to leave it unseeded as a control.  An airplane flew through the cloud in both cases.... [p]recipitation was measured as the total rain volume falling from the cloud base following the airplane seeding run."

```{r, message = FALSE}
clouds <- read_csv("http://www.evanlray.com/data/sleuth3/case0301_cloud_seeding.csv")
head(clouds)
```

#### Starting Point

Here are density plots and box plots, separately for each Treatment.

```{r, fig.height = 2.9}
ggplot(data = clouds, mapping = aes(x = Rainfall, color = Treatment)) +
  geom_density()
```

Standard deviations for each group:

```{r}
clouds %>%
  group_by(Treatment) %>%
  summarize(
    sd_rainfall = sd(Rainfall)
  )
```

Skewed right, so move down one step on the ladder.

\newpage

#### Down 1 Step: $\sqrt{Rainfall}$

```{r}
clouds <- clouds %>%
  mutate(
    sqrt_rainfall = sqrt(Rainfall)
  )
```


```{r, fig.height = 2.9}
ggplot(data = clouds, mapping = aes(x = sqrt_rainfall, color = Treatment)) +
  geom_density()

clouds %>%
  group_by(Treatment) %>%
  summarize(
    sd_rainfall = sd(sqrt_rainfall)
  )
```

Still skewed right, go down another step.

\newpage

#### Down 2 Steps: $\log(Rainfall)$

```{r}
clouds <- clouds %>%
  mutate(
    log_rainfall = log(Rainfall)
  )
```

```{r, fig.height = 2.9}
ggplot(data = clouds, mapping = aes(x = log_rainfall, color = Treatment)) +
  geom_density()

clouds %>%
  group_by(Treatment) %>%
  summarize(
    sd_rainfall = sd(log_rainfall)
  )
```

Good enough!  We can conduct our analysis on this scale.

\newpage

#### Analysis on transformed scale

Separate group means, on log scale

```{r}
clouds %>%
  group_by(Treatment) %>%
  summarize(
    mean_log_rainfall = mean(log_rainfall)
  )
```

Interpret the group mean estimates above on the transformed scale (always works!):

\vspace{3cm}

Interpret the group mean estimates above on the original data scale (works if we got to a place where distributions were approximately symmetric after transformation!):

```{r}
exp(5.13)
exp(3.99)
```

\newpage

```{r}
rainfall_fit <- lm(log_rainfall ~ Treatment, data = clouds)

library(gmodels)
fit.contrast(rainfall_fit, "Treatment", c(1, -1), conf.int = 0.95)
```

Interpret the estimated difference in means above on the transformed scale (always works!):

\vspace{4cm}

Interpret the estimted difference in means above on the original data scale (works only if the transformation selected was the log transformation and the resulting distribution was approximately symmetric!):

```{r}
exp(1.143781)
exp(0.240865)
exp(2.046697)
```

