---
title: "Predictions for \"Simple\" Linear Regression"
subtitle: "(Sleuth 3 Sections 7.4.2 and 7.4.3)"
output: pdf_document
header-includes:
   - \usepackage{soul}
geometry: margin=0.6in
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
library(mosaic)
```

## Goals for today

 * Get an estimate for the mean response at a particular value $x$ of the explanatory variable by plugging in $x$ in the equation
 * Get a *confidence interval* for the mean response at a particular value $x$ using $t$-based methods
 * Make Bonferroni or Scheffe adjustments to get simultaneous confidence intervals for the mean response at multiple values of $x$.

## Example

We have a data set with information about 152 flights by Endeavour Airlines that departed from JFK airport in New York to either Nashville (BNA), Cincinnati (CVG), or Minneapolis-Saint Paul (MSP) in January 2012.

```{r, echo=FALSE, message = FALSE}
flights <- read_csv("flights_data.csv")
```

```{r}
head(flights)
nrow(flights)
```

#### R Code to get model fit

```{r}
model_fit <- lm(air_time ~ distance, data = flights)
summary(model_fit)
```

\newpage

#### Our Estimates for $\beta_0$ and $\beta_1$ are

\vspace{0.5cm}

$\hat{\beta}_0 =$

\vspace{0.5cm}

$\hat{\beta}_1 =$

\vspace{0.5cm}

Note: LaTeX code for $\hat{\beta}_1$ is `\hat{\beta}_1`.

#### Our estimated mean function is

\vspace{0.5cm}

$\hat{\mu}(Y|X) = \hat{\beta}_0 + \hat{\beta}_1 X$ = 

\vspace{0.5cm}

#### The fitted (predicted) mean air time at a flight distance of 589 miles is

\vspace{3cm}

```{r}
14.568 + 0.147 * 589
```

...or...

```{r}
predict_df <- data.frame(
  distance = 589
)
predict_df

predict(model_fit, newdata = predict_df)
```

\newpage

## Standard Error of Estimated Mean

Here's a formula you will never use for the predicted mean at the value $X_0$:

$$SE\{\hat{\mu}(Y|X_0)\} = \hat{\sigma}\sqrt{\frac{1}{n} + \frac{(X_0 - \bar{X})^2}{(n - 1)s_X^2}}$$

 * $\bar{X}$ is the sample mean of the explanatory variable
 * $s^2_X$ is the sample variance of the explanatory variable
 * $\hat{\sigma} = \sqrt{\frac{\text{Sum of Squared Residuals}}{n-2}}$

Something to notice (don't need to memorize):

 * This standard error depends on the value $X_0$ at which we are estimating the mean response
 * $(X_0 - \bar{X})^2$ is smallest if $X_0 = \bar{X}$, so $SE\{\hat{\mu}(Y|X_0)\}$ is smallest at the sample mean.

#### Find and interpret a 95% confidence interval for the mean air time for flights traveling a distance of 589 miles.

```{r}
predict(model_fit, newdata = predict_df, interval = "confidence", se.fit = TRUE)
```

\vspace{5cm}

This is just a $t$-based interval based on the estimate and its standard error (although the calculation of standard error is complicated...)

```{r}
qt(0.975, df = 152 - 2)

101.150 - 1.976 * 1.526
101.150 + 1.976 * 1.526
```

\newpage

#### Find and interpret Bonferroni adjusted confidence intervals for the mean air time at flight distances of 589 miles, 765 miles, and 1029 miles, with a familywise confidence level of 95%.

Approach 1 (easier): adjust confidence level we ask `predict` for.

 * 3 CI's at a familywise confidence level of 95%
 * Overall, miss for 5% of samples, $\alpha = 0.05$
 * Each individual CI has $\alpha = 0.05/3 = 0.0167$
 * Each individual CI has confidence level $(1 - 0.0167) \times 100\% = 98.3\%$

```{r}
predict_df <- data.frame(
  distance = c(589, 765, 1029)
)
predict_df

predict(model_fit,
  newdata = predict_df,
  interval = "confidence",
  se.fit = TRUE,
  level = 0.983
)
```

\newpage

Approach 2: Manual calculation based on standard errors

```{r}
(1 - 0.05/(2*3))
qt(0.9917, df = 152 - 2)
```

```{r}
# CI for X0 = 589
101.150 - 2.423 * 1.526
101.150 + 2.423 * 1.526

# CI for X0 = 765
127.0223 - 2.423 * 0.960
127.0223 + 2.423 * 0.960

# CI for X0 = 1029
165.8301 - 2.423 * 1.016
165.8301 + 2.423 * 1.016
```

#### Find and plot Scheffe adjusted CIs for the means at a grid of 101 values of x between 589 and 1029

```{r, fig.height = 3}
ggplot(data = flights, mapping = aes(x = distance, y = air_time)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()
```

\newpage

#### I will not ask you to do this by hand, the following is just to illustrate

```{r, fig.height=3}
Scheffe_multiplier <- sqrt((2 - 1) * qf(0.95, df1 = 2 - 1, df2 = 152 - 2))

predict_df <- data.frame(
  distance = seq(from = 589, to = 1029, length = 101)
)
head(predict_df)

preds_with_ses <- predict(model_fit,
  newdata = predict_df,
  se.fit = TRUE
)

predict_df <- predict_df %>%
  mutate(
    scheffe_lower = preds_with_ses$fit - Scheffe_multiplier * preds_with_ses$se.fit,
    scheffe_upper = preds_with_ses$fit + Scheffe_multiplier * preds_with_ses$se.fit
  )

ggplot(data = flights, mapping = aes(x = distance, y = air_time)) +
  geom_point() +
  geom_line(data = predict_df, mapping = aes(x = distance, y = scheffe_lower)) +
  geom_line(data = predict_df, mapping = aes(x = distance, y = scheffe_upper)) +
  theme_bw()
```

