---
title: "Anscombe's Data"
output:
  pdf_document:
    keep_tex: true
header-includes:
   - \usepackage{soul}
   - \usepackage{booktabs}
documentclass: extarticle
geometry: margin=0.6in
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
library(mosaic)
library(gridExtra)

anscombe <- read_csv("http://www.evanlray.com/data/base_r/anscombe_quintet.csv")
```

### All 5 Have Essentially the Same Estimated Intercept, Slope, $R^2$, and Residual Standard Deviation!

```{r, echo = FALSE}
fit1 <- lm(y1 ~ x1, data = anscombe)
fit2 <- lm(y2 ~ x2, data = anscombe)
fit3 <- lm(y3 ~ x3, data = anscombe)
fit4 <- lm(y4 ~ x4, data = anscombe)
fit5 <- lm(y5 ~ x5, data = anscombe)

anscombe_long <- data.frame(
  x = c(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4, anscombe$x5),
  y = c(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4, anscombe$y5),
  residual = c(residuals(fit1), residuals(fit2), residuals(fit3), residuals(fit4), residuals(fit5)),
  data_set = rep(paste("Data Set", 1:5), each = nrow(anscombe))
)

grid.arrange(
  ggplot(data = anscombe_long, mapping = aes(x = x, y = y)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~factor(data_set), nrow = 1),
  ggplot(data = anscombe_long, mapping = aes(x = x, y = residual)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~factor(data_set), nrow = 1) +
    ylab("residual\n")
)
```

* Brief reminder of **conditions for linear regression**:
    * **Linear** relationship
    * **Independent** observations
    * **Normally** distributed residuals (or large enough sample size)
    * **Equal variability** of residuals
    * No **outliers**

* How do they check out?
    * Data Set 1 looks OK!
    * Data Set 2 not a linear relationship!
    * Data Sets 3 and 4 have outliers (or influential observations)!
    * Data Set 5 has a lack of equal variance!

* **Conclusion: It's critical to look at plots!!!!!**
