---
title: "Two Lines, Different Slopes (Crocodiles Continued)"
output:
  pdf_document:
    keep_tex: true
header-includes:
   - \usepackage{soul}
   - \usepackage{booktabs}
geometry: margin=0.6in
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
library(mosaic)
require(readr)
```

```{r, message = FALSE, echo = FALSE}
# set seed and sample_frac to get both species in the first few rows.
set.seed(123)

crocs <- read_csv("http://www.evanlray.com/data/sdm4/Crocodile_lengths.csv") %>%
  mutate(Species = factor(Species)) %>%
  sample_frac(size = 1) %>%
  as.data.frame()
```

We have measurements of the head length (cm) and total body length (cm) of 32 crocodiles of two different species:

```{r}
head(crocs, 3)
```

**2 lines by filtering to create separate data sets**

```{r}
aus_crocs <- crocs %>% filter(Species == "Australian")
aus_fit <- lm(Head ~ Body, data = aus_crocs)
summary(aus_fit)
```

```{r}
ind_crocs <- crocs %>% filter(Species == "Indian")
ind_fit <- lm(Head ~ Body, data = ind_crocs)
summary(ind_fit)
```

\newpage

### 1 model, 2 lines (different slopes)

* Last class we used this code to fit 2 lines with the same slope:
```{r, eval = FALSE}
parallel_lines_fit <- lm(Head ~ Body + Species, data = crocs)
```

* To allow for different slopes, use `*` instead of `+`

```{r, echo = TRUE, fig.height=2}
two_lines_fit <- lm(Head ~ Body * Species, data = crocs)
summary(two_lines_fit)

crocs <- crocs %>%
  mutate(
    fitted = predict(two_lines_fit)
  )

ggplot(data = crocs) +
  geom_point(mapping = aes(x = Body, y = Head, color = Species)) +
  geom_line(mapping = aes(x = Body, y = fitted, color = Species))
```

* Recall that `SpeciesIndian` is an indicator variable defined as:
$$\text{{\tt SpeciesIndian}} = \begin{cases} 1 & \text{ if the species for crocodile $i$ is Indian.} \\
0 & \text{ otherwise (in this case, the species is Australian)} \end{cases}
$$

\newpage

#### What is the estimated equation from this model?

\vspace{2.5cm}

#### What is the estimated equation describing the relationship between body length and head length, for Australian crocodiles?

\vspace{2.5cm}

#### What is the estimated equation describing the relationship between body length and head length, for Indian crocodiles?

\vspace{2.5cm}

#### What is the interpretation of $\widehat{\beta}_0 = 3.463$?

\vspace{2.5cm}

#### What is the interpretation of $\widehat{\beta}_1 = 7.075$?

\vspace{2.5cm}

#### What is the interpretation of $\widehat{\beta}_2 = 0.125$?

\vspace{2.5cm}

#### What is the interpretation of $\widehat{\beta}_3 = 0.006$?

\newpage

#### Using the output from the summary function, conduct a test of the claim that the *slope* of the line describing the relationship between body length and head length in the population of all Australian crocodiles is the same as the *slope* of the line describing the relationship between body length and head length in the population of all Indian crocodiles.
