---
title: "2 Way ANOVA (Highlights from Sleuth3 Chapter 13)"
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
  - \usepackage{vwcol}
geometry: margin=0.5in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
library(ggplot2)
library(readr)
library(dplyr)
library(gridExtra)
options(width = 100)
```

**1 way vs 2 way ANOVA**

 * 1 way ANOVA model
    * Quantitative response, categorical explanatory
    * Each group has its own mean, data are normally distributed around the group mean
 * 2 way ANOVA model
    * Quantitative response, 2 categorical explanatory
    * Each combination of groups has its own mean, data are normally distributed around the mean for that combination of groups
 * Actually everything is just a linear model

**Additive Model vs. Interactions Model**

**Additive Model:** `lm(response ~ explanatory1 + explanatory2, data = data)`

 * Group mean = effect from one variable + effect from second variable
 * Visually: Difference in means for levels of variable 1 is constant across levels of variable 2
 * Example: a calibration experiment was performed to explore the relationship between
    * the recorded `counts` from a gas chromatograph (response) -- we use a log transformation to stabilize variance
    * the `concentration` of a compound in a mixture (Low, Medium, or High) and the `flow_rate` through the chromatograph (Slow or Fast)

```{r, echo = FALSE, message = FALSE}
chromatography <- read_csv("http://www.evanlray.com/data/sdm3/Chapter_29/Ch29_Chromatography.csv")
names(chromatography) <- c("concentration", "flow_rate", "counts")
chromatography <- chromatography %>%
  mutate(
    concentration = factor(concentration, levels = c("Low", "Medium", "High"), labels = c("Low", "Medium", "High")),
    log_counts = log(counts)
  )
```

```{r, fig.height = 1.95}
ggplot(data = chromatography, mapping = aes(x = concentration, y = log_counts, color = flow_rate)) +
  geom_boxplot() +
  theme_bw()
```

**Interactions Model:** `lm(response ~ explanatory1 * explanatory2, data = data)`

 * Group mean = specific to combination of levels for variable 1 and 2
 * Visually: Difference in means for levels of variable 1 varies across levels of variable 2
 * Example: A motivated paper airplane thrower measured
    * The `Distance` travelled (response)
    * The `Design` (dart or glider) and whether or not flaps were put on the wings (Flaps or None)

```{r, echo = FALSE, message = FALSE}
planes <- read_csv("~/Documents/teaching/personal_site/elray1.github.io/data/sdm3/Chapter_29/Ch29_Paper_Planes.csv") %>% filter(Paper == "Copier")
```

```{r, fig.height = 1.95}
ggplot(data = planes, mapping = aes(x = flaps, y = Distance, color = Design)) +
  geom_boxplot() +
  theme_bw()
```


\newpage

## Fit group means on transformed scale

```{r}
lm_fit <- lm(log_counts ~ concentration + flow_rate, data = chromatography)
summary(lm_fit)
```

**1. Write down the estimated equation for the mean of `log(counts)` based on concentration and flow rate.**

\vspace{2cm}

**2. Express each of the following estimated means in terms of coefficient estimates above:**

$\hat{\mu}(log(\text{counts}) | \text{concentration = "Low", flow\_rate = "Fast"}) =$

\vspace{0.05cm}

$\hat{\mu}(log(\text{counts}) | \text{concentration = "Medium", flow\_rate = "Fast"}) =$

\vspace{0.05cm}

$\hat{\mu}(log(\text{counts}) | \text{concentration = "High", flow\_rate = "Fast"}) =$

\vspace{0.05cm}

$\hat{\mu}(log(\text{counts}) | \text{concentration = "Low", flow\_rate = "Slow"}) =$

\vspace{0.05cm}

$\hat{\mu}(log(\text{counts}) | \text{concentration = "Medium", flow\_rate = "Slow"}) =$ 

\vspace{0.05cm}

$\hat{\mu}(log(\text{counts}) | \text{concentration = "High", flow\_rate = "Slow"}) =$

**3. Conduct a test of the claim that for a given flow rate, the mean log counts is the same at all three concentrations.**

```{r}
lm_fit_reduced <- lm(log_counts ~ flow_rate, data = chromatography)
anova(lm_fit_reduced, lm_fit)
```

